A weakly structured stem for human origins in Africa

### 2023-01-08 paper.tex
Divergence Discussion

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S -- then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, results in the
formation of an ancestral Khoe-San population $\approx120$ka. The second event
$\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of Eastern/Western Africans including the ancestors of ``Out of
Africans''. The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}.

>>>>

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S -- then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, results in the
formation of an ancestral Khoe-San population $\approx120$ka. The second event
$\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of Eastern/Western Africans including the ancestors of ``Out of
Africans''. While $120$ka can be interpreted as the basal human population divergence,
we suggest concpetualizing this event as the time of most recent shared ancestry among sampled populations.
"Population divergence" in population genetics is a complex concept because of the co-estimation of divergence and migration;
many methods make differing assumptions regarding the degree of a clean split between ancestral groups vs. isolation-with-migration 
(Figure~\ref{fig:supp-IM-misspecification}). Therefore, wide variation exists in estimates of 
divergence time in the literature (e.g. \citep{Pagani2015-pz }).

The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}.
